---
editor: 
  markdown: 
    wrap: 72
---

# Predicting Salary After College

# Corbin Brinkerhoff, Matthew Wilson

## What Should You Really Expect to Earn Five Years After College?

## 1. Introduction

Choosing a college major or stressing over grades can feel like
decisions that will shape the rest of your life, and many students
wonder what these choices actually mean for their future income. To help
answer these questions, we analyzed real data from the American
Community Survey to understand how college graduates are doing five
years after earning their degree.

In this article, we walk through how factors like GPA, major, and gender
relate to early‑career salary. Whether you’re a current student, recent
graduate, or someone advising the next generation, this guide is meant
to give you a clear, data‑driven look at what influences earnings and
what matters less than you might think.

## 2. Impact of GPA

When looking at how grades relate to future earnings, the data shows a
rea, though not overwhelming, connection. On average, each one‑point
increase in GPA is linked to earning about **\$5,400 more per year**,
with the true impact ranging between \$4,700 and \$6,100 for a person.
So yes, GPA does matter, but not in the “your life depends on perfect
grades” way that many students fear. A strong GPA can give you a
meaningful boost, but it’s far from the only thing shaping your salary
after graduation.

Interestingly, the importance of GPA shifts depending on what you study.
While differences across majors aren’t huge, some fields seem to reward
high academic performance more consistently than others. In those
majors, employers may view GPA as a stronger signal of technical skill
or work ethic. In other areas, experience, internships, or a strong
portfolio speak louder than a transcript.

Overall, GPA matters—but it’s just one piece of a much larger puzzle.

## 3. Salaries Across Majors

```{r}
#| echo: false
#| warning: false

library(tidyverse)

sal_df <- vroom::vroom("Salary.csv")

sal_df <- sal_df |>
  mutate(
    MajorCategory = factor(MajorCategory),
    Sex = factor(Sex)
  )

# creating a linear model. Salary predicted my Major, GPA and sex. No interaction
sal_lm<-lm(Salary~ .,sal_df)

# make df for diff majors holding constant of female w/ 3.5 GPA
maj_df <- expand.grid(Sex = "F", GPA = 3.5, MajorCategory = unique(sal_df$MajorCategory))



# Ensure you used the correct arg for base R lm
maj_preds <- predict(sal_lm, newdata = maj_df, se.fit = TRUE)

# Bind predictions back to maj_df
plot_df <- maj_df %>%
  mutate(
    .pred = maj_preds$fit,
  ) %>%
  # Order majors by predicted value for a nicer chart
  mutate(MajorCategory = reorder(MajorCategory, .pred))

library(scales)

ggplot(plot_df, aes(x = MajorCategory, y = .pred)) +
  geom_col(fill = "#4472C4") +
  coord_flip() +  # if you prefer horizontal bars (often easier to read)
  scale_y_continuous(
    breaks = scales::breaks_pretty(n = 5),      # more ticks
    minor_breaks = waiver(),                     # add minor gridlines
    labels = label_number(big.mark = ",")        # or label_dollar() if these are $ values
    # e.g., labels = label_dollar(accuracy = 1)
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor.y = element_line(color = "grey85"),  # emphasize minor gridlines
    panel.grid.major.y = element_line(color = "grey80")
  ) +
  labs(x = "Major category", y = "Predicted Salary", title = "Predicted Salary by Major")
    
```

Our analysis shows clear salary differences across major categories.
Five years after graduating, the highest‑earning fields tend to be:

**Engineering, Computers and Mathematics, and Physical Sciences**

These majors consistently lead to higher predicted salaries. While
salary shouldn’t be the only factor when choosing a field of study, it’s
helpful context for students planning their path toward financial
stability.

## 4. Salaries for Women and Men

When comparing salaries between men and women within the same major
category, we see a persistent difference. Even after accounting for GPA
and field of study, **women earn about \$2,700 less per year** than men
on average.

This gap isn’t identical in every field. Some majors show very small
differences, while others have noticeably larger ones. In fact, the
interaction between gender and major shows that in certain categories,
the salary gap can **widen or narrow** by an **additional**
\$1,000–\$2,300, depending on the nature of the work and industry norms.

Looking within your own major category, these patterns highlight whether
your field tends to reward men and women similarly or whether a
meaningful difference exists. While the exact reasons behind these gaps
are complex, the takeaway is clear: gender‑based salary differences do
exist, and they don’t look the same across all majors.

Understanding these patterns can help students make more informed
decisions and advocate confidently when entering the job market. Here is
a graph that indicates the average difference in salary between men and
women for each major.

```{r}
#| echo: false
#| warning: false

sal_lm_sex_major <- lm(
  Salary ~ GPA + Sex * MajorCategory,
  data = sal_df
)

library(dplyr)
library(tidyr)

# Ensure factors in newdata match the model's training data
# If your model used Sex = c("Female","Male"), adjust the levels/labels accordingly.
maj_df <- expand.grid(
  MajorCategory = levels(sal_df$MajorCategory),
  Sex           = c("F", "M"),
  GPA           = 3.5,
  KEEP.OUT.ATTRS = FALSE,
  stringsAsFactors = FALSE
) %>%
  arrange(MajorCategory, Sex) %>%                 # ensure (Major, then F, then M)
  mutate(
    MajorCategory = factor(MajorCategory, levels = levels(sal_df$MajorCategory)),
    Sex = factor(Sex, levels = levels(sal_df$Sex)) # IMPORTANT: match model levels
  )

# Predict for each row
maj_df$pred <- predict(sal_lm_sex_major, newdata = maj_df)

# Pivot to get columns pred_F and pred_M, then compute Male - Female
major_diffs <- maj_df %>%
  select(MajorCategory, Sex, pred) %>%
  tidyr::pivot_wider(names_from = Sex, values_from = pred, names_prefix = "pred_") %>%
  mutate(diff_M_minus_F = pred_M - pred_F) %>%
  arrange(MajorCategory)

library(ggplot2)
library(dplyr)

# If you want to sort majors by the size of the difference:
major_diffs_plot <- major_diffs %>%
  mutate(
    # order by diff (largest positive at top after coord_flip)
    MajorCategory = reorder(MajorCategory, diff_M_minus_F),
    # color bars by direction of gap
    sign = ifelse(diff_M_minus_F >= 0, "Male > Female", "Female > Male")
  )

ggplot(major_diffs_plot, aes(x = MajorCategory, y = diff_M_minus_F, fill = sign)) +
  geom_col(width = 0.75) +
  geom_hline(yintercept = 0, color = "gray30") +
  coord_flip() +
  scale_fill_manual(values = c("Male > Female" = "#1f77b4", "Female > Male" = "#d62728")) +
  labs(
    title = "Predicted Salary Difference of Male vs Female",
    x = "Major Category",
    y = "Difference (USD, Male − Female)",
    fill = "Direction"
  ) +
  scale_y_continuous(labels = scales::label_dollar()) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = "top"
  )
```

## 5. Other Factors — The Value of a College Education

Our model predicts a graduate’s salary using only three variables—Major,
GPA, and Sex—and on average is off by about \$5,700. That tells us that
while these factors do matter, there’s a lot more influencing income
after college.

Other variables that likely play a role include: Changing industries,
Further education (like a master’s degree), Geographic location, Work
experience and internships, and Personal interest or satisfaction with
the field

To give one example of how our model performs, we randomly selected a
graduate from the dataset: A **female Communications & Journalism
major** with a **GPA of 3.76**. - **Predicted salary:** \$81,694 -
**Actual salary:** \$77,300 - **Difference:** We over‑estimated by
**\$4,394**

This example shows that while our model can capture general trends well,
individual outcomes still vary based on many real‑life factors.

## 6. Conclusions

College decisions like choosing a major or caring about GPA truly do
influence earnings, but they don’t tell the whole story. GPA provides a
helpful salary boost, but it isn’t everything. Your choice of major
matters, with fields like engineering and computer science leading to
higher early‑career earnings. Gender differences in salary persist
across most majors, though the size of the gap varies widely. And
finally, many factors we couldn’t measure like experience, location, and
personal motivation play meaningful roles in shaping future income.

Overall, students should feel empowered, not stressed: your future
salary isn’t determined by any single choice. Instead, it’s shaped by a
combination of your education, experiences, and the opportunities you
pursue along the way.
